---
title: "JUb19 versus CEN2ent"
output:
  html_document:
    df_print: paged
---
```{r load child-JUb19, include=FALSE}
# 1. Extract only the R code from the child Rmd file
temp_script <- tempfile(fileext = ".R")
knitr::purl("3_Jub19_analysis_1.Rmd", output = temp_script, quiet = TRUE)

# 2. Source the extracted code silently into your environment
source(temp_script)
```
Based on May 7th Discussion, JUb19 and CEN2ent became a main interest, and we now conducted the competition experiment under different initial frequencies. This allowed us to estimate competition coefficients via a different approach; thus, we give it a separate treatment here. 

# Data formating from pairwise competition
Compile data from different competition assays (JUb19 as donor and CEN2ent species as recipient) at start (0 hr) and end (24 hr) points (under OD = 1).
```{r, JUbvsCEN-data, echo = FALSE}
JUb19_CEN2ent_raw <- Pairwise_competition_assay_JUb19_vs_CEN2ent_DR_91_19


extract_time_block <- function(raw_df, time_label = "0hr") {
  
  raw_df <- as_tibble(raw_df)
  
  row_has_time <- apply(raw_df, 1, function(x) {
    any(str_squish(as.character(x)) == time_label, na.rm = TRUE)
  })
  
  start_row <- which(row_has_time)[1]
  
  if (is.na(start_row)) {
    stop(paste0("找不到 ", time_label, " 區塊。"))
  }
  
  row_has_any_time <- apply(raw_df, 1, function(x) {
    any(str_squish(as.character(x)) %in% c("0hr", "24hr"), na.rm = TRUE)
  })
  
  next_time_rows <- which(row_has_any_time & seq_len(nrow(raw_df)) > start_row)
  
  end_row <- if (length(next_time_rows) > 0) {
    next_time_rows[1] - 1
  } else {
    nrow(raw_df)
  }
  
  raw_df[start_row:end_row, ] %>%
    filter(!if_all(everything(), ~ is.na(.x) | as.character(.x) == ""))
}

make_pairwise_ratio_long <- function(raw_df,
                                     donor_name = "JUb19",
                                     recipient_colname = "Recipient: CEN2ent",
                                     growth_time = "0hr") {
  
  parse_num2 <- function(x) {
    x <- as.character(x)
    x <- str_replace_all(x, ",", "")
    x <- str_replace_all(x, "\\s+", "")
    suppressWarnings(as.numeric(x))
  }
  
  log10_or_na <- function(x) {
    ifelse(!is.na(x) & x > 0, log10(x), NA_real_)
  }
  
  raw_df <- as_tibble(raw_df)
  raw_names <- names(raw_df)
  
  start_idx <- which(str_detect(raw_names, regex("^Donor\\s*:", ignore_case = TRUE)))
  
  if (length(start_idx) == 0) {
    stop("找不到 Donor: 開頭的欄位，請確認 Excel 第一列是否有被讀成欄名。")
  }
  
  blocks <- map(seq_along(start_idx), function(i) {
    
    start <- start_idx[i]
    end <- if (i < length(start_idx)) start_idx[i + 1] - 1 else ncol(raw_df)
    
    block_title <- raw_names[start]
    
    ratio <- str_match(
      block_title,
      regex("D\\s*:\\s*R\\s*=\\s*([^;]+)", ignore_case = TRUE)
    )[, 2]
    
    ratio <- str_squish(ratio)
    
    recipient_from_title <- str_match(
      block_title,
      regex("Recipient\\s*:\\s*([^;]+)", ignore_case = TRUE)
    )[, 2]
    
    recipient_from_title <- str_squish(recipient_from_title)
    
    block <- raw_df[, start:end]
    
    block <- block[, !map_lgl(block, ~ all(is.na(.x) | as.character(.x) == "")), drop = FALSE]
    
    if (ncol(block) < 2) {
      return(NULL)
    }
    
    new_names <- as.character(unlist(block[1, ]))
    
    new_names[1] <- "metric"
    
    new_names[-1] <- ifelse(
      is.na(new_names[-1]) | new_names[-1] == "",
      recipient_from_title,
      new_names[-1]
    )
    
    new_names <- make.unique(new_names)
    colnames(block) <- new_names
    
    block <- block[-1, ]
    
    block <- block %>%
      filter(!if_all(everything(), ~ is.na(.x) | as.character(.x) == ""))
    
    block_long <- block %>%
      pivot_longer(
        cols = -metric,
        names_to = "strain_raw",
        values_to = "value"
      ) %>%
      mutate(
        metric = str_squish(as.character(metric)),
        ratio = ratio,
        growth_time = growth_time,
        
        replicate = case_when(
          str_detect(strain_raw, "\\.\\d+$") ~ as.integer(str_extract(strain_raw, "\\d+$")) + 1L,
          TRUE ~ 1L
        ),
        
        strain = strain_raw %>%
          str_remove("\\.\\.\\.\\d+$") %>%
          str_remove("\\.\\d+$")
      ) %>%
      select(strain, replicate, ratio, growth_time, metric, value)
    
    return(block_long)
  })
  
  tidy_df <- bind_rows(blocks)
  
  pairwise_df <- tidy_df %>%
    mutate(
      value = parse_num2(value),
      
      metric = case_when(
        str_detect(metric, regex("^Donor\\s*:", ignore_case = TRUE)) ~ paste0("Donor: ", donor_name),
        str_detect(metric, regex("^Recipient\\s*:", ignore_case = TRUE)) ~ recipient_colname,
        TRUE ~ metric
      )
    ) %>%
    
    group_by(strain, replicate, ratio, growth_time) %>%
    mutate(
      cfu_order = cumsum(metric == "CFU/ml"),
      metric = case_when(
        metric == "CFU/ml" & cfu_order == 1 ~ "Donor_CFU",
        metric == "CFU/ml" & cfu_order == 2 ~ "Recipient_CFU",
        TRUE ~ metric
      )
    ) %>%
    ungroup() %>%
    select(-cfu_order) %>%
    
    pivot_wider(
      id_cols = c(strain, replicate, ratio, growth_time),
      names_from = metric,
      values_from = value,
      values_fn = first
    ) %>%
    
    mutate(
      `log donor` = log10_or_na(Donor_CFU),
      `log recipient` = log10_or_na(Recipient_CFU)
    ) %>%
    
    select(
      strain,
      replicate,
      ratio,
      Donor_CFU,
      Recipient_CFU,
      `log donor`,
      `log recipient`,
      growth_time
    )
  
  return(pairwise_df)
}

## 1. 從同一個 sheet 裡切出 0hr 和 24hr
JUb19_CEN2ent_0hr_block <- extract_time_block(
  raw_df = JUb19_CEN2ent_raw,
  time_label = "0hr"
)

JUb19_CEN2ent_24hr_block <- extract_time_block(
  raw_df = JUb19_CEN2ent_raw,
  time_label = "24hr"
)

## 2. 分別整理成 long dataset
JUb19_CEN2ent_0hr <- make_pairwise_ratio_long(
  raw_df = JUb19_CEN2ent_0hr_block,
  donor_name = "JUb19",
  recipient_colname = "Recipient: CEN2ent",
  growth_time = "0hr"
)

JUb19_CEN2ent_24hr <- make_pairwise_ratio_long(
  raw_df = JUb19_CEN2ent_24hr_block,
  donor_name = "JUb19",
  recipient_colname = "Recipient: CEN2ent",
  growth_time = "24hr"
)

## 3. 合併成你要的 final dataset
JUb19_CEN2ent_DR_91_19 <- inner_join(
  JUb19_CEN2ent_0hr %>%
    select(
      strain,
      replicate,
      ratio,
      D0 = Donor_CFU,
      R0 = Recipient_CFU,
      logD0 = `log donor`,
      logR0 = `log recipient`
    ),
  
  JUb19_CEN2ent_24hr %>%
    select(
      strain,
      replicate,
      ratio,
      D24 = Donor_CFU,
      R24 = Recipient_CFU,
      logD24 = `log donor`,
      logR24 = `log recipient`
    ),
  
  by = c("strain", "replicate", "ratio")
) %>%
  select(
    strain,
    replicate,
    D0,
    R0,
    logD0,
    logR0,
    D24,
    R24,
    logD24,
    logR24,
    ratio
  ) %>%
  arrange(strain, ratio, replicate)

# JUb19_CEN2ent_DR_91_19

## 檢查 1:9 / 9:1 有沒有都進來
# JUb19_CEN2ent_DR_91_19 %>%
#   count(ratio)

## 分開看
JUb19_CEN2ent_DR_91_19_9to1 <- JUb19_CEN2ent_DR_91_19 %>%
  filter(ratio == "9:1")

JUb19_CEN2ent_DR_91_19_1to9 <- JUb19_CEN2ent_DR_91_19 %>%
  filter(ratio == "1:9")

JUb19_CEN2ent_DR_55 <- pairJUb19 %>%
  filter(strain == "CEN2ent") %>%
  mutate(ratio = "5:5")

JUb19_CEN2ent_all_ratio <- bind_rows(
  JUb19_CEN2ent_DR_55,
  JUb19_CEN2ent_DR_91_19
)

JUb19_CEN2ent_all_ratio

```


# Frequency lineplot (JUb19 vs CEN2ent)
Here, as a first attempt to visualize the competitive outcome, we plot the frequency change across 0 and 24 hrs. The first figure shows the frequency change for JUb19 (donor) and the second mirrored figure shows that for CEN2ent (recipient). Overall, there's a signature of species coexistence: the rare species increase in frequency whereas the abundant species decreases in frequency.
```{r JUBvsCEN-lineplot, echo = FALSE}
# 計算頻率
Donor_frequency_df <- JUb19_CEN2ent_all_ratio %>%
    mutate(
        D_0hr = D0 / (D0 + R0),
        D_24hr = D24 / (D24 + R24)
    ) %>%
    pivot_longer(cols = c(D_0hr, D_24hr), names_to = "Time", values_to = "Frequency")

Recipient_frequency_df <- JUb19_CEN2ent_all_ratio %>%
    mutate(
        R_0hr = R0 / (D0 + R0),
        R_24hr = R24 / (D24 + R24)
    ) %>%
    pivot_longer(cols = c(R_0hr, R_24hr), names_to = "Time", values_to = "Frequency")

# 畫斜率圖 (Slope graph)
# donor JUb19
ggplot(Donor_frequency_df, 
       aes(x = Time, y = Frequency, group = interaction(strain, replicate, ratio), color = strain)) +
    geom_line(alpha = 0.5) +
    geom_point() +
    facet_wrap(~strain) +
  scale_color_manual(values = my_colors)+
    theme_minimal() +
    labs(title = "Change in JUb19 Frequency compare to CEN2ent (0hr vs 24hr)",
          x = "Time")

# recipient others
ggplot(Recipient_frequency_df, 
       aes(x = Time, y = Frequency, group = interaction(strain, replicate, ratio), color = strain)) +
    geom_line(alpha = 0.5) +
    geom_point() +
    facet_wrap(~strain) +
  scale_color_manual(values = my_colors)+
    theme_minimal() +
    labs(title = "Change in CEN2ent Frequency compare to JUb19 (0hr vs 24hr)",
          x = "Time")
```

# Interspecies coefficient alpha (JUb19 vs CEN2ent) 
Again, consider a Lotkta--Volterra competition model in the form of $\frac{dN_i}{dt} = N_i(r_i-\alpha_{ii}N_i-\alpha_{ij}N_j)$. Previously, we estimated the interspecific competition coefficients ($\alpha_{ij}$) by rearranging the growth expression. That is:
\[\alpha_{ij} = \left[\hat{r_i}-\hat{\alpha_{ii}}N_{i0}^c-\left(\frac{lnN_{i24}^c - lnN_{i0}^c}{24})\right)\right].\left(\frac{1}{N_{j0}^c}\right)\] 
Here, however, since we have a large initial density gradient for the competition between JUb19 and CEN2ent, we can apply the same statistical fitting approach used to obtained the intraspecific competition coefficients. We did this is a "sequential" approach: the response variable is the difference between the observed per capita growth rate ($\frac{\Delta lnN_i}{\Delta t}$) and the intraspecific competitive impact (i.e., $\hat{r_i}-\hat{\alpha_{ii}}N_{i0}^c$), termed the "residual" in the following figures. This residual term is modeled as a function of the initial heterospecific competitor density ($N_{j0}^c$), and the resulting slope represents the interspecific competition coefficient. Note again all density data were scaled by $10^9$ prior to calculation. Below, we show (left) the estimated effect of CEN2ent on JUb19 ($\alpha_{\textrm{JUb19, CEN2ent}}$), and (right) the estimated effect of JUb19 on CEN2ent ($\alpha_{\textrm{CEN2ent, JUb19}}$). A more detailed comparison of this approach and the previous one, as well as the simulated over competitive outcome, will follow soon. 
```{r JUBvsCEN-coefficient, echo = FALSE}
# ============================================================
# JUb19 vs CEN2ent statistical fit for beta_DR and beta_RD
# Parameter source: pooled Nstart monoculture model
# ============================================================

scale_factor <- 1e9


# ============================================================
# 1. Prepare monoculture-derived parameters
# ============================================================

mono_param_pooled_Nstart_beta <- selflimit_pooled_N_start %>%
  ungroup() %>%
  transmute(
    strain,
    r_i,
    alpha_ii_signed,
    beta_ii_selflimit = -alpha_ii_signed
  )

param_JUb19 <- mono_param_pooled_Nstart_beta %>%
  filter(strain == "JUb19") %>%
  slice(1)

param_CEN2ent <- mono_param_pooled_Nstart_beta %>%
  filter(strain == "CEN2ent") %>%
  slice(1)
# 
# param_JUb19
# param_CEN2ent


# ============================================================
# 2. Prepare JUb19-CEN2ent pairwise fitting dataframe
# D = JUb19
# R = CEN2ent
# ============================================================

JUb19_CEN2ent_fit_df <- JUb19_CEN2ent_all_ratio %>%
  mutate(
    ratio = factor(ratio, levels = c("1:9", "5:5", "9:1")),
    
    D0 = as.numeric(D0)/ scale_factor,
    R0 = as.numeric(R0)/ scale_factor,
    D24 = as.numeric(D24)/ scale_factor,
    R24 = as.numeric(R24)/ scale_factor,
    
    D0_s = D0 ,
    R0_s = R0 ,
    
    g_D = log(D24 / D0) / 24,
    g_R = log(R24 / R0) / 24,
    
    # Donor = JUb19 parameters
    r_D = param_JUb19$r_i,
    beta_DD = param_JUb19$beta_ii_selflimit,
    
    # Recipient = CEN2ent parameters
    r_R = param_CEN2ent$r_i,
    beta_RR = param_CEN2ent$beta_ii_selflimit,
    
     # Rearranged model:
    # y_DR = beta_DR * R0_s
    # y_RD = beta_RD * D0_s
    y_DR = r_D - beta_DD * D0_s - g_D,
    y_RD = r_R - beta_RR * R0_s - g_R,
    
    # replicate-wise finite-difference estimate
    beta_DR_obs = y_DR / R0_s,
    beta_RD_obs = y_RD / D0_s,
    
    # signed alpha version
    alpha_DR_signed_obs = -beta_DR_obs,
    alpha_RD_signed_obs = -beta_RD_obs
  )

# JUb19_CEN2ent_fit_df

# ============================================================
# Overall fit
# beta_DR: CEN2ent effect on JUb19
# beta_RD: JUb19 effect on CEN2ent
# ============================================================

fit_beta_DR_overall <- lm(y_DR  ~ R0_s, data = JUb19_CEN2ent_fit_df)
fit_beta_RD_overall <- lm(y_RD  ~ D0_s, data = JUb19_CEN2ent_fit_df)

# summary(fit_beta_DR_overall)
# summary(fit_beta_RD_overall)

overall_beta_summary <- bind_rows(
  broom::tidy(fit_beta_DR_overall) %>%
    mutate(
      interaction = "alpha_DR: CEN2ent effect on JUb19",
      beta_estimate = estimate,
      alpha_signed = -estimate
    ),
  
  broom::tidy(fit_beta_RD_overall) %>%
    mutate(
      interaction = "alpha_RD: JUb19 effect on CEN2ent",
      beta_estimate = estimate,
      alpha_signed = -estimate
    )
) %>%
  select(
    interaction,
    term,
    beta_estimate,
    std.error,
    statistic,
    p.value,
    alpha_signed
  ) %>%
  mutate(
    interpretation = case_when(
      beta_estimate > 0 & p.value < 0.05 ~ "significant suppression",
      beta_estimate > 0 ~ "suppression trend",
      beta_estimate < 0 & p.value < 0.05 ~ "significant facilitation-like effect",
      beta_estimate < 0 ~ "facilitation-like trend",
      TRUE ~ "unclear"
    )
  )

# overall_beta_summary
# ============================================================
# Ratio-specific fit
# Estimate beta_DR and beta_RD separately for each ratio
# ============================================================

# fit_one_beta <- function(df, response_var, predictor_var) {
#   
#   formula_use <- as.formula(
#     paste(response_var, "~ ", predictor_var)
#   )
#   
#   fit <- lm(formula_use, data = df)
#   
#   broom::tidy(fit) %>%
#     select(
#       term,
#       estimate,
#       std.error,
#       statistic,
#       p.value
#     )
# }

# ratio_beta_DR_summary <- JUb19_CEN2ent_fit_df %>%
#   group_by(ratio) %>%
#   nest() %>%
#   mutate(
#     fit_result = map(data, ~ fit_one_beta(.x, "y_DR", "R0_s"))
#   ) %>%
#   select(ratio, fit_result) %>%
#   unnest(fit_result) %>%
#   mutate(
#     interaction = "alpha_DR: CEN2ent effect on JUb19",
#     beta_estimate = estimate,
#     alpha_signed = -estimate
#   )
# 
# ratio_beta_RD_summary <- JUb19_CEN2ent_fit_df %>%
#   group_by(ratio) %>%
#   nest() %>%
#   mutate(
#     fit_result = map(data, ~ fit_one_beta(.x, "y_RD", "D0_s"))
#   ) %>%
#   select(ratio, fit_result) %>%
#   unnest(fit_result) %>%
#   mutate(
#     interaction = "alpha_RD: JUb19 effect on CEN2ent",
#     beta_estimate = estimate,
#     alpha_signed = -estimate
#   )
# 
# ratio_beta_summary <- bind_rows(
#   ratio_beta_DR_summary,
#   ratio_beta_RD_summary
# ) %>%
#   select(
#     ratio,
#     interaction,
#     term,
#     beta_estimate,
#     std.error,
#     statistic,
#     p.value,
#     alpha_signed
#   ) %>%
#   mutate(
#     interpretation = case_when(
#       beta_estimate > 0 & p.value < 0.05 ~ "significant suppression",
#       beta_estimate > 0 ~ "suppression trend",
#       beta_estimate < 0 & p.value < 0.05 ~ "significant facilitation-like effect",
#       beta_estimate < 0 ~ "facilitation-like trend",
#       TRUE ~ "unclear"
#     )
#   )
# ratio_beta_summary


# JUb19_CEN2ent_beta_obs_summary <- JUb19_CEN2ent_fit_df %>%
#   group_by(ratio) %>%
#   summarise(
#     n = n(),
#     
#     mean_beta_DR_obs = mean(beta_DR_obs, na.rm = TRUE),
#     sd_beta_DR_obs = sd(beta_DR_obs, na.rm = TRUE),
#     se_beta_DR_obs = sd_beta_DR_obs / sqrt(n),
#     
#     mean_beta_RD_obs = mean(beta_RD_obs, na.rm = TRUE),
#     sd_beta_RD_obs = sd(beta_RD_obs, na.rm = TRUE),
#     se_beta_RD_obs = sd_beta_RD_obs / sqrt(n),
#     
#     .groups = "drop"
#   )
# JUb19_CEN2ent_beta_obs_summary

overall_beta_summary_intercept <- bind_rows(
  
  broom::tidy(fit_beta_DR_overall) %>%
    mutate(
      interaction = "alpha_DR: CEN2ent effect on JUb19",
      model = "with intercept"
    ),
  
  broom::tidy(fit_beta_RD_overall) %>%
    mutate(
      interaction = "alpha_RD: JUb19 effect on CEN2ent",
      model = "with intercept"
    )
) %>%
  select(
    interaction,
    model,
    term,
    estimate,
    std.error,
    statistic,
    p.value
  )

overall_beta_summary_wide <- overall_beta_summary_intercept %>%
  mutate(
    parameter = case_when(
      term == "(Intercept)" ~ "intercept",
      term %in% c("R0_s", "D0_s") ~ "beta",
      TRUE ~ term
    )
  ) %>%
  select(
    interaction,
    model,
    parameter,
    estimate,
    std.error,
    statistic,
    p.value
  ) %>%
  pivot_wider(
    names_from = parameter,
    values_from = c(estimate, std.error, statistic, p.value)
  ) %>%
  mutate(
    alpha_signed = -estimate_beta,
    estimate_intercept_signed_form = -estimate_intercept,
    interpretation = case_when(
      estimate_beta > 0 & p.value_beta < 0.05 ~ "significant suppression",
      estimate_beta > 0 ~ "suppression trend",
      estimate_beta < 0 & p.value_beta < 0.05 ~ "significant facilitation-like effect",
      estimate_beta < 0 ~ "facilitation-like trend",
      TRUE ~ "unclear"
    )
  )
beta_fit_plot_df <- JUb19_CEN2ent_fit_df %>%
  select(
    ratio,
    replicate,
    R0_s,
    D0_s,
    y_DR,
    y_RD,
    alpha_DR_signed_obs,
    alpha_RD_signed_obs
  ) %>%
  mutate(signed_y_DR = -y_DR,
         signed_y_RD = -y_RD) %>%
  pivot_longer(
    cols = c(signed_y_DR, signed_y_RD),   # change y value
    names_to = "response_type",
    values_to = "y_value"
  ) %>%
  mutate(
    x_value = case_when(                  # change x value
      response_type == "signed_y_DR" ~ R0_s,
      response_type == "signed_y_RD" ~ D0_s
    ),
    interaction = case_when(
      response_type == "signed_y_DR" ~ "alpha_DR: CEN2ent effect on JUb19",
      response_type == "signed_y_RD" ~ "alpha_RD: JUb19 effect on CEN2ent"
    )
  )

interaction_slope_plot <- ggplot(
  beta_fit_plot_df,
  aes(
    x = x_value,
    y = y_value
  )
) +
  geom_point(
    aes(color = interaction),
    size = 2.5,
    alpha = 0.75
  ) +
  geom_abline(
    data = overall_beta_summary_wide,
    aes(
      slope = alpha_signed, intercept = estimate_intercept_signed_form,
      color = interaction
    ),
    linewidth = 1
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "red"
  ) +
  facet_wrap(~ interaction, scales = "free_x") +
  theme_bw() +
  labs(
    title = "Statistical fitting of interspecific limitation coefficients",
    subtitle = "JUb19 as donor, CEN2ent as recipient",
    x = "Scaled competitor initial density",
    y = "Residual growth limitation term",
    color = "Interaction"
  ) +
  theme(
    aspect.ratio = 0.75
  )

print(interaction_slope_plot)
```
