---
title: "JUb19 as donor versus all other recipients"
output:
  html_document:
    df_print: paged
---
```{r load child-JUb19, include=FALSE}
# 1. Extract only the R code from the child Rmd file
temp_script <- tempfile(fileext = ".R")
knitr::purl("3_Jub19_analysis_1.Rmd", output = temp_script, quiet = TRUE)

# 2. Source the extracted code silently into your environment
source(temp_script)
```

# Data formating from pairwise competition
Compile data from different competition assays (MSPm1 as donor and other species as recipient) at start (0 hr) and end (24 hr) points (under OD = 1).
```{r MSPm1-data, echo = FALSE}
pairMSPm1_0hr_untreated <- Pairwise_culture_MSPm1_0hr

colnames(pairMSPm1_0hr_untreated) <- pairMSPm1_0hr_untreated[1,]
pairMSPm1_0hr_untreated <- pairMSPm1_0hr_untreated[2:35,]
# 1. find block starts
start_idx <- c(0, which(pairMSPm1_0hr_untreated$`0hr` == "0hr")) 

# 2. extract blocks
blocks <- map(seq_along(start_idx), function(i) {
  
  start <- start_idx[i]
  end <- if (i < length(start_idx)) start_idx[i + 1] - 1 else nrow(pairMSPm1_0hr_untreated)
  
  block <- pairMSPm1_0hr_untreated[start:end, ]
  colnames(block) <- make.unique(colnames(block))
  # remove empty rows
  block <- block %>% filter(!if_all(everything(), is.na))
  
  if (i == 1) {
    # FIRST BLOCK → KEEP ORIGINAL COLUMN NAMES
    block <- block %>%
      rename(metric = `0hr`) #
    
  } else {
    # OTHER BLOCKS → USE HEADER ROW
    
    new_names <- as.character(unlist(block[1, ]))
    

# remove NA column names completely
    valid_cols <- !is.na(new_names) & new_names != ""
    
    block <- block[, valid_cols]
    new_names <- new_names[valid_cols]
    
# make names unique (important!)
new_names <- make.unique(new_names)

colnames(block) <- new_names
  
  # remove header row
  block <- block[-1, ]
  
  # rename first column
  names(block)[1] <- "metric"
  }
  # reshape long
  block_long <- block %>%
    pivot_longer(-metric, names_to = "strain", values_to = "value")
  
  return(block_long)
})

# 3. combine
tidy_df <- bind_rows(blocks)

# 4. clean numbers
tidy_df <- tidy_df %>%
  mutate(value = parse_number(value))

# 5. assign replicate IDs
pairMSPm1_0hr <- tidy_df %>%
  group_by(strain) %>%
  mutate(metric = case_when(
    metric == "CFU/ml" & row_number() < 5 ~ "Donor_CFU",     # 前半段的 CFU
    metric == "CFU/ml" & row_number() >= 5 ~ "Recipient_CFU", # 後半段的 CFU
    TRUE ~ metric
  )) %>%
  ungroup() %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  
  # FIRST clean strain names
  mutate(
    strain = str_remove(strain, "\\.\\.\\.\\d+$"),
    strain = str_remove(strain, "\\.\\d+$")
  ) %>%
  
  # THEN assign replicate correctly
  group_by(strain) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  
  mutate(`growth time` = "0hr") %>%  #
  select(strain, replicate, everything())

# extract JUb19 to pairMSPm1
extracted_JUb19_data <- pairJUb19_0hr %>%
  filter(strain == "MSPm1") %>%
  mutate(
    strain = "JUb19", 
    temp_D_CFU = Recipient_CFU, 
    temp_R_CFU = Donor_CFU,
    temp_logD = `log recipient`,
    temp_logR = `log donor`
  ) %>%
  mutate(
    Donor_CFU = temp_D_CFU,
    Recipient_CFU = temp_R_CFU,
    `log donor` = temp_logD,
    `log recipient` = temp_logR
  ) %>%
  select(-temp_D_CFU, -temp_R_CFU, -temp_logD, -temp_logR) # 移除暫存變數

pairMSPm1_0hr <- bind_rows(pairMSPm1_0hr, extracted_JUb19_data) %>%
  arrange(strain, replicate)
# pairMSPm1_0hr

## 24 hr
pairMSPm1_24hr_untreated <- Pairwise_culture_MSPm1_24hr

colnames(pairMSPm1_24hr_untreated) <- pairMSPm1_24hr_untreated[1,]
pairMSPm1_24hr_untreated <- pairMSPm1_24hr_untreated[2:35,]
# 1. find block starts
start_idx <- c(0, which(pairMSPm1_24hr_untreated$`24hr` == "24hr")) 

# 2. extract blocks
blocks <- map(seq_along(start_idx), function(i) {
  
  start <- start_idx[i]
  end <- if (i < length(start_idx)) start_idx[i + 1] - 1 else nrow(pairMSPm1_24hr_untreated)
  
  block <- pairMSPm1_24hr_untreated[start:end, ]
  colnames(block) <- make.unique(colnames(block))
  # remove empty rows
  block <- block %>% filter(!if_all(everything(), is.na))
  
  if (i == 1) {
    # FIRST BLOCK → KEEP ORIGINAL COLUMN NAMES
    block <- block %>%
      rename(metric = `24hr`) #
    
  } else {
    # OTHER BLOCKS → USE HEADER ROW
    
    new_names <- as.character(unlist(block[1, ]))
    

# remove NA column names completely
    valid_cols <- !is.na(new_names) & new_names != ""
    
    block <- block[, valid_cols]
    new_names <- new_names[valid_cols]
    
# make names unique (important!)
new_names <- make.unique(new_names)

colnames(block) <- new_names
  
  # remove header row
  block <- block[-1, ]
  
  # rename first column
  names(block)[1] <- "metric"
  }
  # reshape long
  block_long <- block %>%
    pivot_longer(-metric, names_to = "strain", values_to = "value")
  
  return(block_long)
})

# 3. combine
tidy_df <- bind_rows(blocks)

# 4. clean numbers
tidy_df <- tidy_df %>%
  mutate(value = parse_number(value))

# 5. assign replicate IDs
pairMSPm1_24hr <- tidy_df %>%
  group_by(strain) %>%
  mutate(metric = case_when(
    metric == "CFU/ml" & row_number() < 5 ~ "Donor_CFU",     # 前半段的 CFU
    metric == "CFU/ml" & row_number() >= 5 ~ "Recipient_CFU", # 後半段的 CFU
    TRUE ~ metric
  )) %>%
  ungroup() %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  
  # FIRST clean strain names
  mutate(
    strain = str_remove(strain, "\\.\\.\\.\\d+$"),
    strain = str_remove(strain, "\\.\\d+$")
  ) %>%
  
  # THEN assign replicate correctly
  group_by(strain) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  
  mutate(`growth time` = "24hr") %>%  #
  select(strain, replicate, everything())

# extract JUb19 to pairMSPm1
extracted_JUb19_data <- pairJUb19_24hr %>%
  filter(strain == "MSPm1") %>%
  mutate(
    strain = "JUb19", 
    temp_D_CFU = Recipient_CFU, 
    temp_R_CFU = Donor_CFU,
    temp_logD = `log recipient`,
    temp_logR = `log donor`
  ) %>%
  mutate(
    Donor_CFU = temp_D_CFU,
    Recipient_CFU = temp_R_CFU,
    `log donor` = temp_logD,
    `log recipient` = temp_logR
  ) %>%
  select(-temp_D_CFU, -temp_R_CFU, -temp_logD, -temp_logR) 

pairMSPm1_24hr <- bind_rows(pairMSPm1_24hr, extracted_JUb19_data) %>%
  arrange(strain, replicate)
# pairMSPm1_24hr

pairMSPm1 <- inner_join(
  pairMSPm1_0hr %>% select(strain, replicate, D0 = Donor_CFU, R0 = Recipient_CFU, 
                           logD0 ='log donor', logR0 ='log recipient'),
  pairMSPm1_24hr %>% select(strain, replicate, D24 = Donor_CFU, R24 = Recipient_CFU,
                            logD24 ='log donor', logR24 ='log recipient'),
  by = c("strain", "replicate")
)
pairMSPm1
```

# Frequency lineplot (MSPm1 vs others)
Here, as a first attempt to visualize the competitive outcome, we plot the frequency change across 0 and 24 hrs. The first figure shows the frequency change for MSPm1 (donor) and the second mirrored figure shows that for the other species (recipient). As expected, MSPm1 is indeed much less superior compared to the previous competition experiments involving JUb19. The frequency of MSPm1 plummets when competing with JUb19 and MYb10.  
```{r MSPm1-lineplot, echo = FALSE}
# 計算頻率
Donor_frequency_df <- pairMSPm1 %>%
    mutate(
        D_0hr = D0 / (D0 + R0),
        D_24hr = D24 / (D24 + R24)
    ) %>%
    pivot_longer(cols = c(D_0hr, D_24hr), names_to = "Time", values_to = "Frequency")

Recipient_frequency_df <- pairMSPm1 %>%
    mutate(
        R_0hr = R0 / (D0 + R0),
        R_24hr = R24 / (D24 + R24)
    ) %>%
    pivot_longer(cols = c(R_0hr, R_24hr), names_to = "Time", values_to = "Frequency")

# 畫斜率圖 (Slope graph)
# donor MSPm1
ggplot(Donor_frequency_df, 
       aes(x = Time, y = Frequency, group = interaction(strain, replicate), color = strain)) +
    geom_line(alpha = 0.5) +
    geom_point() +
    facet_wrap(~strain) +
  scale_color_manual(values = my_colors)+
    theme_minimal() +
    labs(title = "Change in MSPm1 Frequency (0hr vs 24hr)",
          x = "Time")

# reciepient others 
ggplot(Recipient_frequency_df, 
       aes(x = Time, y = Frequency, group = interaction(strain, replicate), color = strain)) +
    geom_line(alpha = 0.5) +
    geom_point() +
    facet_wrap(~strain) +
  scale_color_manual(values = my_colors)+
    theme_minimal() +
    labs(title = "Change in Recipient Frequency compare to MSPm1(0hr vs 24hr)",
          x = "Time")
```

<!--
# Bivariate fitness plots (MSPm1 vs others)
```{r MSPm1-bivariate, echo = FALSE}
log_plot_df <- pairMSPm1 %>%
    mutate(
        log_growth_donor = logD24 - logD0,       # Donor 的成長量 (D24-D0)/D0
        log_growth_recipient = logR24 - logR0    # Recipient 的成長量
    )



## bivariate fitness plot individual
Biplot_indi_MSPm1 <- ggplot(log_plot_df, aes(x = log_growth_donor, y = log_growth_recipient, color = strain)) + 
    # 加上 1:1 的等量成長線 (紅虛線)
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 0.8) +
    geom_hline(yintercept = 0) +
    # 畫出 6 個 Replicate 的散佈點
    geom_point(size = 3, alpha = 0.7) +
    # 加上每個 Strain 的平均值中心點 (更大的點)
    stat_summary(fun = mean, geom = "point", size = 6, shape = 18) +
    # 套用 13 色固定色盤
    scale_color_manual(values = my_colors) +
    theme_bw() +
    labs(
        title = "MSPm1 Log10 Fitness Plot: Donor vs Recipient",
        x = "Donor Log10 Growth (0 -> 24hr)",
        y = "Recipient Log10 Growth (0 -> 24hr)",
        color = "Bacterial Strain"
    ) 

print(Biplot_indi_MSPm1)
```
--> 

# Interspecies competition value between donor and recepient (MSPm1 vs others) 
Again, consider a Lotkta--Volterra competition model in the form of $\frac{dN_i}{dt} = N_i(r_i-\alpha_{ii}N_i-\alpha_{ij}N_j)$. While we used a statistical fitting approach to estimate the intraspecific competition coefficients ($\alpha_{ii}$H), we cannot do that for the interspecific competition coefficients ($\alpha_{ij}$) due to the lack of a density gradient (but see the results for Jub19 and CEN2ent in a separate file). Instead, we use a "hybrid" approach. Here, in the Lotka--Volterra competition model, the per capita growth rate is still estimated as $\frac{\Delta lnN_i}{\Delta t}$, and the species' $r_i$ and $\alpha_{ii}$ are obtained by the previous statistical fitting results for the monoculture growth assay (i.e., $\hat{r_i}$ and $\hat{\alpha_{ii}}$). By rearranging the equation, we can indirectly obtain an esitmate for the interspecific competition coefficients as follows:   
\[\alpha_{DR} = \left[\hat{r_D}-\hat{\alpha_{DD}}N_{D0}^c-\left(\frac{lnN_{D24}^c - lnN_{D0}^c}{24}\right)\right].\left(\frac{1}{N_{R0}^c}\right)\]
\[\alpha_{RD} = \left[\hat{r_R}-\hat{\alpha_{RR}}N_{R0}^c-\left(\frac{lnN_{R24}^c - lnN_{R0}^c}{24}\right)\right].\left(\frac{1}{N_{D0}^c}\right)\], 
where $N_{jt}^c$ represents the scaled population size of species $j$ (= $D$ of $R$) at time $t$ (= 0 or 24 hr) within the co-culture (superscript $c$). Note again all density data were scaled by $10^9$ prior to calculation. Below, we show (left) the estimated effect of other recipient species of MSPm1 ($\alpha_{DR}$), and (right) the estimated effect of MSPm1 on the other recipient species ($\alpha_{RD}$, with $D$ = MSPm1). Interestingly, the effect of MSPm1 on both JUb19 and MYb10 are weak compared to reciprocal interactions, which may explain why MSPm1 plummets when competed against these two species.
```{r MSPm1-coefficient, echo = FALSE}
scale_factor <- 1e9

# ============================================================
# 1. Prepare monoculture-derived parameters
# Use pooled Nstart model
# ============================================================

mono_param_pooled_Nstart_beta <- selflimit_pooled_N_start %>%
  ungroup() %>%
  transmute(
    strain,
    r_i,
    alpha_ii_signed,
    
    # Positive self-limitation coefficient
    # g = r_i - beta_ii_selflimit * Nstart_s
    beta_ii_selflimit = -alpha_ii_signed
  )

# Donor = MSPm1
param_D <- mono_param_pooled_Nstart_beta %>%
  filter(strain == "MSPm1") %>%
  slice(1)

# param_D


# ============================================================
# 2. Calculate beta_DR and beta_RD
# D = Donor = MSPm1
# R = Recipient = CeMbio strain
# ============================================================

fitness_df_MSPm1_pooled_Nstart <- pairMSPm1 %>%
  left_join(
    mono_param_pooled_Nstart_beta %>%
      rename(
        r_R = r_i,
        alpha_RR_signed = alpha_ii_signed,
        beta_RR_selflimit = beta_ii_selflimit
      ),
    by = "strain"
  ) %>%
  mutate(
    # -----------------------------
    # Donor = MSPm1
    # -----------------------------
    N_D_0_comp = as.numeric(D0)/scale_factor,
    N_D_24_comp = as.numeric(D24)/scale_factor,
    
    # -----------------------------
    # Recipient = CeMbio strain
    # -----------------------------
    N_R_0_comp = as.numeric(R0)/scale_factor,
    N_R_24_comp = as.numeric(R24)/scale_factor,
    
    # -----------------------------
    # Scaled starting density
    # Must match pooled Nstart model
    # -----------------------------
    D0_s = N_D_0_comp ,
    R0_s = N_R_0_comp ,
    
    # -----------------------------
    # Observed pairwise growth rate
    # -----------------------------
    g_D = log(N_D_24_comp / N_D_0_comp) / 24,
    g_R = log(N_R_24_comp / N_R_0_comp) / 24,
    
    # -----------------------------
    # Donor monoculture parameters
    # from pooled Nstart model
    # -----------------------------
    r_D = param_D$r_i,
    alpha_DD_signed = param_D$alpha_ii_signed,
    beta_DD_selflimit = param_D$beta_ii_selflimit,
    
    # ========================================================
    # Positive limitation model
    #
    # g_D = r_D - beta_DD * D0_s - beta_DR * R0_s
    # g_R = r_R - beta_RR * R0_s - beta_RD * D0_s
    # ========================================================
    
    # beta_DR:
    # Recipient effect on donor MSPm1
    # beta_DR > 0 means recipient suppresses MSPm1
    beta_DR_pooled_Nstart =
      (r_D - beta_DD_selflimit * D0_s - g_D) / R0_s,
    
    # beta_RD:
    # Donor MSPm1 effect on recipient
    # beta_RD > 0 means MSPm1 suppresses recipient
    beta_RD_pooled_Nstart =
      (r_R - beta_RR_selflimit * R0_s - g_R) / D0_s,
    
    # Signed alpha version
    # alpha < 0 means suppression
    alpha_DR_signed_pooled_Nstart = -beta_DR_pooled_Nstart,
    alpha_RD_signed_pooled_Nstart = -beta_RD_pooled_Nstart
  )

# fitness_df_MSPm1_pooled_Nstart
fitness_summary_MSPm1_pooled_Nstart <- fitness_df_MSPm1_pooled_Nstart %>%
  group_by(strain) %>%
  summarise(
    n = n(),
    
    mean_beta_DR = mean(beta_DR_pooled_Nstart, na.rm = TRUE),
    sd_beta_DR = sd(beta_DR_pooled_Nstart, na.rm = TRUE),
    se_beta_DR = sd_beta_DR / sqrt(sum(!is.na(beta_DR_pooled_Nstart))),
    
    mean_beta_RD = mean(beta_RD_pooled_Nstart, na.rm = TRUE),
    sd_beta_RD = sd(beta_RD_pooled_Nstart, na.rm = TRUE),
    se_beta_RD = sd_beta_RD / sqrt(sum(!is.na(beta_RD_pooled_Nstart))),
    
    mean_alpha_DR_signed = mean(alpha_DR_signed_pooled_Nstart, na.rm = TRUE),
    sd_alpha_DR_signed = sd(alpha_DR_signed_pooled_Nstart, na.rm = TRUE),
    
    mean_alpha_RD_signed = mean(alpha_RD_signed_pooled_Nstart, na.rm = TRUE),
    sd_alpha_RD_signed = sd(alpha_RD_signed_pooled_Nstart, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  mutate(
    DR_interpretation = case_when(
      mean_beta_DR > 0 ~ "Recipient suppresses MSPm1",
      mean_beta_DR < 0 ~ "Recipient facilitates MSPm1",
      TRUE ~ "neutral"
    ),
    RD_interpretation = case_when(
      mean_beta_RD > 0 ~ "MSPm1 suppresses recipient",
      mean_beta_RD < 0 ~ "MSPm1 facilitates recipient",
      TRUE ~ "neutral"
    )
  )

# fitness_summary_MSPm1_pooled_Nstart
fitness_beta_long_MSPm1 <- fitness_df_MSPm1_pooled_Nstart %>%
  select(
    strain,
    replicate,
    beta_DR_pooled_Nstart,
    beta_RD_pooled_Nstart
  ) %>%
  pivot_longer(
    cols = c(beta_DR_pooled_Nstart, beta_RD_pooled_Nstart),
    names_to = "beta_type",
    values_to = "beta_value"
  ) %>%
  mutate(
    beta_type = case_when(
      beta_type == "beta_DR_pooled_Nstart" ~
        "alpha_DR: Recipient effect on MSPm1",
      beta_type == "beta_RD_pooled_Nstart" ~
        "alpha_RD: MSPm1 effect on Recipient",
      TRUE ~ beta_type
    )
  )

interComp_beta_plot_MSPm1_pooled_Nstart <- ggplot(
  fitness_beta_long_MSPm1,
  aes(
    x = reorder(strain, beta_value),
    y = beta_value,
    color = strain
  )
) +
  geom_jitter(
    width = 0.15,
    alpha = 0.5,
    size = 2
  ) +
  stat_summary(
    fun.data = "mean_se",
    geom = "errorbar",
    width = 0.3,
    size = 1,
    color = "black"
  ) +
  stat_summary(
    fun = "mean",
    geom = "point",
    size = 4,
    color = "black",
    shape = 18
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "red"
  ) +
  scale_color_manual(values = my_colors) +
  facet_wrap(~ beta_type, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Interspecies competition index statistical fitting",
    subtitle = "MSPm1 as donor",
    x = "Bacterial recipient",
    y = "Interspecific limitation coefficient alpha"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

print(interComp_beta_plot_MSPm1_pooled_Nstart)
```

