
---
title: "C_elegans_project_AS(frequency_slope, growth_scatter, alpha_donor_recipient"
output: html_notebook
---
#libraries
```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(knitr)
library(ggplot2)
library(tidyverse)
library(deSolve)
library(RColorBrewer)
library(Hmisc)
```

#load data (change pathway here)
```{r}
## Monoculture
mono_culture_growth_assay_of_CeMbio_0hr <- read_excel("D:/ntu/11_kekelab/cElegent_project_AS/20250809 mono-culture growth assay of CeMbio strains wo nitrocellulose membrane.xlsx", 
    sheet = "0 hr")
mono_culture_growth_assay_of_CeMbio_24hr <- read_excel("D:/ntu/11_kekelab/cElegent_project_AS/20250809 mono-culture growth assay of CeMbio strains wo nitrocellulose membrane.xlsx", 
    sheet = "24 hr")

##Pairwise culture
#Jub19 vs other
Pairwise_culture_JUb19_0hr <- read_excel("D:/ntu/11_kekelab/cElegent_project_AS/20260111 Pairwise competition assay (JUb19 vs CeMbio)-summary.xlsx", 
    sheet = "0 hr")
Pairwise_culture_JUb19_24hr <- read_excel("D:/ntu/11_kekelab/cElegent_project_AS/20260111 Pairwise competition assay (JUb19 vs CeMbio)-summary.xlsx", 
    sheet = "24 hr")

#pair wise culture MSPm1 vs other (lack Jub19)
Pairwise_culture_MSPm1_0hr <- read_excel("D:/ntu/11_kekelab/cElegent_project_AS/20260414 Pairwise competition assay (MSPm1 vs CeMbio)-summary.xlsx", 
    sheet = "0 hr")
Pairwise_culture_MSPm1_24hr <- read_excel("D:/ntu/11_kekelab/cElegent_project_AS/20260414 Pairwise competition assay (MSPm1 vs CeMbio)-summary.xlsx", 
    sheet = "24 hr")


JUb19_male_leaving_raw_data <- read_excel("D:/ntu/11_kekelab/cElegent_project_AS/JUb19 male leaving raw data.xlsx", 
    sheet = "Sheet1")

Pairwise_competition_assay_JUb19_vs_CEN2ent_DR_91_19 <- read_excel("D:/ntu/11_kekelab/cElegent_project_AS/20260508 Pairwise competition assay (JUb19 vs CEN2ent)-DR=91 ; 19.xlsx")

```

## Data format reconstruction
#Data formating from monoculture (mono_0hr, mono_24hr, mono_df(conbine 0hr and 24hr), monoMean_df(calculate strains mean of mono_df))
```{r}
# Monoculture 0 hr treatment
mono_0hr_untreat <- mono_culture_growth_assay_of_CeMbio_0hr

# 1. find block starts
start_idx <- c(0, which(mono_0hr_untreat$`0hr` == "0hr"))

# 2. extract blocks
blocks <- map(seq_along(start_idx), function(i) {
  
  start <- start_idx[i]
  end <- if (i < length(start_idx)) start_idx[i + 1] - 1 else nrow(mono_0hr_untreat)
  
  block <- mono_0hr_untreat[start:end, ]
  
  # remove empty rows
  block <- block %>% filter(!if_all(everything(), is.na))
  
  if (i == 1) {
    # FIRST BLOCK → KEEP ORIGINAL COLUMN NAMES
    block <- block %>%
      rename(metric = `0hr`)
    
  } else {
    new_names <- as.character(unlist(block[1, ]))
    

# remove NA column names completely
    valid_cols <- !is.na(new_names) & new_names != ""
    
    block <- block[, valid_cols]
    new_names <- new_names[valid_cols]
    
# make names unique (important!)
new_names <- make.unique(new_names)

colnames(block) <- new_names
  
  # remove header row
  block <- block[-1, ]
  
  # rename first column
  names(block)[1] <- "metric"
  }
  # reshape long
  block_long <- block %>%
    pivot_longer(-metric, names_to = "strain", values_to = "value")
  
  return(block_long)
})

# 3. combine
tidy_df <- bind_rows(blocks)

# 4. clean numbers
tidy_df <- tidy_df %>%
  mutate(value = parse_number(value))

# 5. assign replicate IDs
mono_0hr <- tidy_df %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  
  # FIRST clean strain names
  mutate(
    strain = str_remove(strain, "\\.\\.\\.\\d+$"),
    strain = str_remove(strain, "\\.\\d+$")
  ) %>%
  
  # THEN assign replicate correctly
  group_by(strain) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  
  mutate(`growth time` = "0hr") %>%  #
  select(strain, replicate, everything())

mono_0hr

## Monoculture 24 hr treatment
mono_24hr_untreat <- mono_culture_growth_assay_of_CeMbio_24hr

# 1. find block starts
start_idx <- c(0, which(mono_24hr_untreat$`24hr` == "24hr"))

# 2. extract blocks
blocks <- map(seq_along(start_idx), function(i) {
  
  start <- start_idx[i]
  end <- if (i < length(start_idx)) start_idx[i + 1] - 1 else nrow(mono_24hr_untreat)
  
  block <- mono_24hr_untreat[start:end, ]
  
  # remove empty rows
  block <- block %>% filter(!if_all(everything(), is.na))
  
  if (i == 1) {
    # FIRST BLOCK → KEEP ORIGINAL COLUMN NAMES
    block <- block %>%
      rename(metric = `24hr`)
    
  } else {
    # OTHER BLOCKS → USE HEADER ROW
    
    new_names <- as.character(unlist(block[1, ]))
    

# remove NA column names completely
    valid_cols <- !is.na(new_names) & new_names != ""
    
    block <- block[, valid_cols]
    new_names <- new_names[valid_cols]
    
# make names unique (important!)
new_names <- make.unique(new_names)

colnames(block) <- new_names
  
  # remove header row
  block <- block[-1, ]
  
  # rename first column
  names(block)[1] <- "metric"
  }
  # reshape long
  block_long <- block %>%
    pivot_longer(-metric, names_to = "strain", values_to = "value")
  
  return(block_long)
})

# 3. combine
tidy_df <- bind_rows(blocks)

# 4. clean numbers
tidy_df <- tidy_df %>%
  mutate(value = parse_number(value))

# 5. assign replicate IDs
mono_24hr <- tidy_df %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  
  # FIRST clean strain names
  mutate(
    strain = str_remove(strain, "\\.\\.\\.\\d+$"),
    strain = str_remove(strain, "\\.\\d+$")
  ) %>%
  
  # THEN assign replicate correctly
  group_by(strain) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  
  mutate(`growth time` = "24hr") %>%  #
  select(strain, replicate, everything())

mono_24hr

## combine mono_0hr and mono_24hr
mono_df <- inner_join(
  mono_0hr %>% select(strain, replicate, M0 = `CFU/ml`,logM0 = `log CFU/ml`),
  mono_24hr %>% select(strain, replicate, M24 = `CFU/ml`,logM24 = `log CFU/ml`),
  by = c("strain", "replicate")
)
mono_df

monoMean_df <- mono_df %>%
  group_by(strain) %>%
  summarise(mean_M0 = mean(M0, na.rm = TRUE), mean_M24=mean(M24, na.rm = TRUE), 
            mean_logM0 = mean(logM0, na.rm = TRUE), mean_logM24 = mean(logM24, na.rm = TRUE), 
            .groups = "drop")

monoMean_df
```

## Data formating from pairwise competition 
#JUb19
```{r}
## 0 hr
pairJUb19_0hr_untreated <- Pairwise_culture_JUb19_0hr

colnames(pairJUb19_0hr_untreated) <- pairJUb19_0hr_untreated[1,]
pairJUb19_0hr_untreated <- pairJUb19_0hr_untreated[2:35,]
# 1. find block starts
start_idx <- c(0, which(pairJUb19_0hr_untreated$`0hr` == "0hr")) 

# 2. extract blocks
blocks <- map(seq_along(start_idx), function(i) {
  start <- start_idx[i]
  end <- if (i < length(start_idx)) start_idx[i + 1] - 1 else nrow(pairJUb19_0hr_untreated)
  
  block <- pairJUb19_0hr_untreated[start:end, ]
  colnames(block) <- make.unique(colnames(block))
  # remove empty rows
  block <- block %>% filter(!if_all(everything(), is.na))
  
  if (i == 1) {
    # FIRST BLOCK → KEEP ORIGINAL COLUMN NAMES
    block <- block %>%
      rename(metric = `0hr`) #
  } else {
    # OTHER BLOCKS → USE HEADER ROW
    new_names <- as.character(unlist(block[1, ]))
    

# remove NA column names completely
    valid_cols <- !is.na(new_names) & new_names != ""
    
    block <- block[, valid_cols]
    new_names <- new_names[valid_cols]
    
# make names unique (important!)
new_names <- make.unique(new_names)

colnames(block) <- new_names
  # remove header row
  block <- block[-1, ]
  # rename first column
  names(block)[1] <- "metric"
  }
  # reshape long
  block_long <- block %>%
    pivot_longer(-metric, names_to = "strain", values_to = "value")
  return(block_long)
})

# 3. combine
tidy_df <- bind_rows(blocks)

# 4. clean numbers
tidy_df <- tidy_df %>%
  mutate(value = parse_number(value))

# 5. assign replicate IDs
pairJUb19_0hr <- tidy_df %>%
  group_by(strain) %>%
  mutate(metric = case_when(
    metric == "CFU/ml" & row_number() < 5 ~ "Donor_CFU",     # 前半段的 CFU
    metric == "CFU/ml" & row_number() >= 5 ~ "Recipient_CFU", # 後半段的 CFU
    TRUE ~ metric
  )) %>%
  ungroup() %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  
  # FIRST clean strain names
  mutate(
    strain = str_remove(strain, "\\.\\.\\.\\d+$"),
    strain = str_remove(strain, "\\.\\d+$")
  ) %>%
  
  # THEN assign replicate correctly
  group_by(strain) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  
  mutate(`growth time` = "0hr") %>%  #
  select(strain, replicate, everything())

pairJUb19_0hr

##  24 hr
pairJUb19_24hr_untreated <- Pairwise_culture_JUb19_24hr

colnames(pairJUb19_24hr_untreated) <- pairJUb19_24hr_untreated[1,]
pairJUb19_24hr_untreated <- pairJUb19_24hr_untreated[2:35,]
# 1. find block starts
start_idx <- c(0, which(pairJUb19_24hr_untreated$`24hr` == "24hr")) 

# 2. extract blocks
blocks <- map(seq_along(start_idx), function(i) {
  
  start <- start_idx[i]
  end <- if (i < length(start_idx)) start_idx[i + 1] - 1 else nrow(pairJUb19_24hr_untreated)
  
  block <- pairJUb19_24hr_untreated[start:end, ]
  colnames(block) <- make.unique(colnames(block))
  # remove empty rows
  block <- block %>% filter(!if_all(everything(), is.na))
  
  if (i == 1) {
    # FIRST BLOCK → KEEP ORIGINAL COLUMN NAMES
    block <- block %>%
      rename(metric = `24hr`) #
    
  } else {
    # OTHER BLOCKS → USE HEADER ROW
    
    new_names <- as.character(unlist(block[1, ]))
    

# remove NA column names completely
    valid_cols <- !is.na(new_names) & new_names != ""
    
    block <- block[, valid_cols]
    new_names <- new_names[valid_cols]
    
# make names unique (important!)
new_names <- make.unique(new_names)

colnames(block) <- new_names
  
  # remove header row
  block <- block[-1, ]
  
  # rename first column
  names(block)[1] <- "metric"
  }
  # reshape long
  block_long <- block %>%
    pivot_longer(-metric, names_to = "strain", values_to = "value")
  
  return(block_long)
})

# 3. combine
tidy_df <- bind_rows(blocks)

# 4. clean numbers
tidy_df <- tidy_df %>%
  mutate(value = parse_number(value))

# 5. assign replicate IDs
pairJUb19_24hr <- tidy_df %>%
  group_by(strain) %>%
  mutate(metric = case_when(
    metric == "CFU/ml" & row_number() < 5 ~ "Donor_CFU",     # 前半段的 CFU
    metric == "CFU/ml" & row_number() >= 5 ~ "Recipient_CFU", # 後半段的 CFU
    TRUE ~ metric
  )) %>%
  ungroup() %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  
  # FIRST clean strain names
  mutate(
    strain = str_remove(strain, "\\.\\.\\.\\d+$"),
    strain = str_remove(strain, "\\.\\d+$")
  ) %>%
  
  # THEN assign replicate correctly
  group_by(strain) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  
  mutate(`growth time` = "0hr") %>%  #
  select(strain, replicate, everything())
pairJUb19_24hr

pairJUb19 <- inner_join(
  pairJUb19_0hr %>% select(strain, replicate, D0 = Donor_CFU, R0 = Recipient_CFU, 
                           logD0 ='log donor', logR0 ='log recipient'),
  pairJUb19_24hr %>% select(strain, replicate, D24 = Donor_CFU, R24 = Recipient_CFU,
                            logD24 ='log donor', logR24 ='log recipient'),
  by = c("strain", "replicate")
)
pairJUb19 <- pairJUb19 %>%
  filter(strain != "JUb19 (control)")
pairJUb19
```
#MSPm1
```{r}
## 0 hr
pairMSPm1_0hr_untreated <- Pairwise_culture_MSPm1_0hr

colnames(pairMSPm1_0hr_untreated) <- pairMSPm1_0hr_untreated[1,]
pairMSPm1_0hr_untreated <- pairMSPm1_0hr_untreated[2:35,]
# 1. find block starts
start_idx <- c(0, which(pairMSPm1_0hr_untreated$`0hr` == "0hr")) 

# 2. extract blocks
blocks <- map(seq_along(start_idx), function(i) {
  
  start <- start_idx[i]
  end <- if (i < length(start_idx)) start_idx[i + 1] - 1 else nrow(pairMSPm1_0hr_untreated)
  
  block <- pairMSPm1_0hr_untreated[start:end, ]
  colnames(block) <- make.unique(colnames(block))
  # remove empty rows
  block <- block %>% filter(!if_all(everything(), is.na))
  
  if (i == 1) {
    # FIRST BLOCK → KEEP ORIGINAL COLUMN NAMES
    block <- block %>%
      rename(metric = `0hr`) #
    
  } else {
    # OTHER BLOCKS → USE HEADER ROW
    
    new_names <- as.character(unlist(block[1, ]))
    

# remove NA column names completely
    valid_cols <- !is.na(new_names) & new_names != ""
    
    block <- block[, valid_cols]
    new_names <- new_names[valid_cols]
    
# make names unique (important!)
new_names <- make.unique(new_names)

colnames(block) <- new_names
  
  # remove header row
  block <- block[-1, ]
  
  # rename first column
  names(block)[1] <- "metric"
  }
  # reshape long
  block_long <- block %>%
    pivot_longer(-metric, names_to = "strain", values_to = "value")
  
  return(block_long)
})

# 3. combine
tidy_df <- bind_rows(blocks)

# 4. clean numbers
tidy_df <- tidy_df %>%
  mutate(value = parse_number(value))

# 5. assign replicate IDs
pairMSPm1_0hr <- tidy_df %>%
  group_by(strain) %>%
  mutate(metric = case_when(
    metric == "CFU/ml" & row_number() < 5 ~ "Donor_CFU",     # 前半段的 CFU
    metric == "CFU/ml" & row_number() >= 5 ~ "Recipient_CFU", # 後半段的 CFU
    TRUE ~ metric
  )) %>%
  ungroup() %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  
  # FIRST clean strain names
  mutate(
    strain = str_remove(strain, "\\.\\.\\.\\d+$"),
    strain = str_remove(strain, "\\.\\d+$")
  ) %>%
  
  # THEN assign replicate correctly
  group_by(strain) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  
  mutate(`growth time` = "0hr") %>%  #
  select(strain, replicate, everything())

# extract JUb19 to pairMSPm1
extracted_JUb19_data <- pairJUb19_0hr %>%
  filter(strain == "MSPm1") %>%
  mutate(
    strain = "JUb19", 
    temp_D_CFU = Recipient_CFU, 
    temp_R_CFU = Donor_CFU,
    temp_logD = `log recipient`,
    temp_logR = `log donor`
  ) %>%
  mutate(
    Donor_CFU = temp_D_CFU,
    Recipient_CFU = temp_R_CFU,
    `log donor` = temp_logD,
    `log recipient` = temp_logR
  ) %>%
  select(-temp_D_CFU, -temp_R_CFU, -temp_logD, -temp_logR) # 移除暫存變數

pairMSPm1_0hr <- bind_rows(pairMSPm1_0hr, extracted_JUb19_data) %>%
  arrange(strain, replicate)

pairMSPm1_0hr

## 24 hr
pairMSPm1_24hr_untreated <- Pairwise_culture_MSPm1_24hr

colnames(pairMSPm1_24hr_untreated) <- pairMSPm1_24hr_untreated[1,]
pairMSPm1_24hr_untreated <- pairMSPm1_24hr_untreated[2:35,]
# 1. find block starts
start_idx <- c(0, which(pairMSPm1_24hr_untreated$`24hr` == "24hr")) 

# 2. extract blocks
blocks <- map(seq_along(start_idx), function(i) {
  
  start <- start_idx[i]
  end <- if (i < length(start_idx)) start_idx[i + 1] - 1 else nrow(pairMSPm1_24hr_untreated)
  
  block <- pairMSPm1_24hr_untreated[start:end, ]
  colnames(block) <- make.unique(colnames(block))
  # remove empty rows
  block <- block %>% filter(!if_all(everything(), is.na))
  
  if (i == 1) {
    # FIRST BLOCK → KEEP ORIGINAL COLUMN NAMES
    block <- block %>%
      rename(metric = `24hr`) #
    
  } else {
    # OTHER BLOCKS → USE HEADER ROW
    
    new_names <- as.character(unlist(block[1, ]))
    

# remove NA column names completely
    valid_cols <- !is.na(new_names) & new_names != ""
    
    block <- block[, valid_cols]
    new_names <- new_names[valid_cols]
    
# make names unique (important!)
new_names <- make.unique(new_names)

colnames(block) <- new_names
  
  # remove header row
  block <- block[-1, ]
  
  # rename first column
  names(block)[1] <- "metric"
  }
  # reshape long
  block_long <- block %>%
    pivot_longer(-metric, names_to = "strain", values_to = "value")
  
  return(block_long)
})

# 3. combine
tidy_df <- bind_rows(blocks)

# 4. clean numbers
tidy_df <- tidy_df %>%
  mutate(value = parse_number(value))

# 5. assign replicate IDs
pairMSPm1_24hr <- tidy_df %>%
  group_by(strain) %>%
  mutate(metric = case_when(
    metric == "CFU/ml" & row_number() < 5 ~ "Donor_CFU",     # 前半段的 CFU
    metric == "CFU/ml" & row_number() >= 5 ~ "Recipient_CFU", # 後半段的 CFU
    TRUE ~ metric
  )) %>%
  ungroup() %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  
  # FIRST clean strain names
  mutate(
    strain = str_remove(strain, "\\.\\.\\.\\d+$"),
    strain = str_remove(strain, "\\.\\d+$")
  ) %>%
  
  # THEN assign replicate correctly
  group_by(strain) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  
  mutate(`growth time` = "24hr") %>%  #
  select(strain, replicate, everything())

# extract JUb19 to pairMSPm1
extracted_JUb19_data <- pairJUb19_24hr %>%
  filter(strain == "MSPm1") %>%
  mutate(
    strain = "JUb19", 
    temp_D_CFU = Recipient_CFU, 
    temp_R_CFU = Donor_CFU,
    temp_logD = `log recipient`,
    temp_logR = `log donor`
  ) %>%
  mutate(
    Donor_CFU = temp_D_CFU,
    Recipient_CFU = temp_R_CFU,
    `log donor` = temp_logD,
    `log recipient` = temp_logR
  ) %>%
  select(-temp_D_CFU, -temp_R_CFU, -temp_logD, -temp_logR) 

pairMSPm1_24hr <- bind_rows(pairMSPm1_24hr, extracted_JUb19_data) %>%
  arrange(strain, replicate)

pairMSPm1_24hr

pairMSPm1 <- inner_join(
  pairMSPm1_0hr %>% select(strain, replicate, D0 = Donor_CFU, R0 = Recipient_CFU, 
                           logD0 ='log donor', logR0 ='log recipient'),
  pairMSPm1_24hr %>% select(strain, replicate, D24 = Donor_CFU, R24 = Recipient_CFU,
                            logD24 ='log donor', logR24 ='log recipient'),
  by = c("strain", "replicate")
)
pairMSPm1
```
## color selection for plots
```{r}
all_strains <- sort(unique(c(mono_24hr$strain, mono_24hr$strain)))
base_colors <- c(brewer.pal(12, "Paired"), brewer.pal(8, "Set2"))
get_palette <- colorRampPalette(base_colors)
my_colors <- setNames(get_palette(length(all_strains)), all_strains)
```

## Frequency lineplot
#JUb19 vs others
```{r}
# 計算頻率
Donor_frequency_df <- pairJUb19 %>%
    mutate(
        D_0hr = D0 / (D0 + R0),
        D_24hr = D24 / (D24 + R24)
    ) %>%
    pivot_longer(cols = c(D_0hr, D_24hr), names_to = "Time", values_to = "Frequency")

Recipient_frequency_df <- pairJUb19 %>%
    mutate(
        R_0hr = R0 / (D0 + R0),
        R_24hr = R24 / (D24 + R24)
    ) %>%
    pivot_longer(cols = c(R_0hr, R_24hr), names_to = "Time", values_to = "Frequency")

# 畫斜率圖 (Slope graph)
# donor JUb19
ggplot(Donor_frequency_df, 
       aes(x = Time, y = Frequency, group = interaction(strain, replicate), color = strain)) +
    geom_line(alpha = 0.5) +
    geom_point() +
    facet_wrap(~strain) +
  scale_color_manual(values = my_colors)+
    theme_minimal() +
    labs(title = "Change in JUb19 Frequency (0hr vs 24hr)",
          x = "Time")

# recipient others
ggplot(Recipient_frequency_df, 
       aes(x = Time, y = Frequency, group = interaction(strain, replicate), color = strain)) +
    geom_line(alpha = 0.5) +
    geom_point() +
    facet_wrap(~strain) +
  scale_color_manual(values = my_colors)+
    theme_minimal() +
    labs(title = "Change in Recipient Frequency compare to JUb19(0hr vs 24hr)",
          x = "Time")
```

#MSPm1 vs others
```{r}
# 計算頻率
Donor_frequency_df <- pairMSPm1 %>%
    mutate(
        D_0hr = D0 / (D0 + R0),
        D_24hr = D24 / (D24 + R24)
    ) %>%
    pivot_longer(cols = c(D_0hr, D_24hr), names_to = "Time", values_to = "Frequency")

Recipient_frequency_df <- pairMSPm1 %>%
    mutate(
        R_0hr = R0 / (D0 + R0),
        R_24hr = R24 / (D24 + R24)
    ) %>%
    pivot_longer(cols = c(R_0hr, R_24hr), names_to = "Time", values_to = "Frequency")

# 畫斜率圖 (Slope graph)
# donor MSPm1
ggplot(Donor_frequency_df, 
       aes(x = Time, y = Frequency, group = interaction(strain, replicate), color = strain)) +
    geom_line(alpha = 0.5) +
    geom_point() +
    facet_wrap(~strain) +
  scale_color_manual(values = my_colors)+
    theme_minimal() +
    labs(title = "Change in MSPm1 Frequency (0hr vs 24hr)",
          x = "Time")

# reciepient others 
ggplot(Recipient_frequency_df, 
       aes(x = Time, y = Frequency, group = interaction(strain, replicate), color = strain)) +
    geom_line(alpha = 0.5) +
    geom_point() +
    facet_wrap(~strain) +
  scale_color_manual(values = my_colors)+
    theme_minimal() +
    labs(title = "Change in Recipient Frequency compare to MSPm1(0hr vs 24hr)",
          x = "Time")
```


## Bivariate fitness plots
# JUb19
```{r}
log_plot_df <- pairJUb19 %>%
    mutate(
        log_growth_donor = logD24 - logD0,       # Donor 的成長量 (D24-D0)/D0
        log_growth_recipient = logR24 - logR0    # Recipient 的成長量
    )



## bivariate fitness plot individual
Biplot_indi_JUb19 <- ggplot(log_plot_df, aes(x = log_growth_donor, y = log_growth_recipient, color = strain)) + 
    # 加上 1:1 的等量成長線 (紅虛線)
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 0.8) +
    geom_hline(yintercept = 0) +
    # 畫出 6 個 Replicate 的散佈點
    geom_point(size = 3, alpha = 0.7) +
    # 加上每個 Strain 的平均值中心點 (更大的點)
    stat_summary(fun = mean, geom = "point", size = 6, shape = 18) +
    # 套用 13 色固定色盤
    scale_color_manual(values = my_colors) +
    theme_bw() +
    labs(
        title = "JUb19 Log10 Fitness Plot: Donor vs Recipient",
        x = "Donor Log10 Growth (0 -> 24hr)",
        y = "Recipient Log10 Growth (0 -> 24hr)",
        color = "Bacterial Strain"
    ) 

print(Biplot_indi_JUb19)
```

# MSPm1
```{r}
log_plot_df <- pairMSPm1 %>%
    mutate(
        log_growth_donor = logD24 - logD0,       # Donor 的成長量 (D24-D0)/D0
        log_growth_recipient = logR24 - logR0    # Recipient 的成長量
    )



## bivariate fitness plot individual
Biplot_indi_MSPm1 <- ggplot(log_plot_df, aes(x = log_growth_donor, y = log_growth_recipient, color = strain)) + 
    # 加上 1:1 的等量成長線 (紅虛線)
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 0.8) +
    geom_hline(yintercept = 0) +
    # 畫出 6 個 Replicate 的散佈點
    geom_point(size = 3, alpha = 0.7) +
    # 加上每個 Strain 的平均值中心點 (更大的點)
    stat_summary(fun = mean, geom = "point", size = 6, shape = 18) +
    # 套用 13 色固定色盤
    scale_color_manual(values = my_colors) +
    theme_bw() +
    labs(
        title = "MSPm1 Log10 Fitness Plot: Donor vs Recipient",
        x = "Donor Log10 Growth (0 -> 24hr)",
        y = "Recipient Log10 Growth (0 -> 24hr)",
        color = "Bacterial Strain"
    ) 

print(Biplot_indi_MSPm1)
```


## interspecies competition value(alpha ) between donor and recepient
#JUb19
```{r}

# 提取 Donor (JUb19) 的單獨成長平均值 (Ki)
K_Donor_val <- monoMean_df %>%
  filter(strain == "JUb19") %>%
  summarise(K_D_mean = mean_M24, r_D = (mean_logM24 - mean_logM0)/24 )

fitness_df_JUb19 <- pairJUb19 %>%
  # 合併 Recipient 的 K 值
  left_join(monoMean_df, by = "strain") %>%
  mutate(
    # 定義 Ki (Donor) 為剛才算出的常數
    N_D_24_mono = K_Donor_val$K_D_mean,
    N_D_24_comp = (D24),
    N_D_0_comp = D0,
    N_R_0_comp = (R0) ,
    r_D = K_Donor_val$r_D,
    # 計算 interspecies competition
    interComp = ((N_D_24_comp-N_D_24_mono)/24)*(-1/N_R_0_comp)
  )

fitness_df_JUb19

interComp_plot_JUb19 <- ggplot(fitness_df_JUb19, 
                               aes(x = reorder(strain, interComp), y = interComp, color = strain)) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.3, size = 1, color = "black") +
  stat_summary(fun = "mean", geom = "point", size = 4, color = "black", shape = 18) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  scale_color_manual(values = my_colors)+
  theme_minimal() +
  labs(title = "Interspecies competition (Mean ± SE)",
       subtitle = "JUb19 as donor",
       x = "Bacterial recepient", y = "Interspecies competition") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

print(interComp_plot_JUb19)

mono_JUb19_raw <- mono_df %>%
  filter(strain == "JUb19") %>%
  # 確認欄位沒有 NA，並計算出每次重複的 r_D
  mutate(r_D_raw = (logM24 - logM0) / 24)

# 設定 Bootstrap 抽樣次數
n_boot <- 1000

# 步驟 2：執行 Bootstrap 重抽樣
set.seed(42)
boot_fitness_df <- pairJUb19 %>%
  group_by(strain) %>%
  group_modify(~ {
    pair_data <- .x
    
    # 隨機抽出 1000 次 Mono 的索引值 (replace = TRUE 允許重複抽樣)
    mono_idx <- sample(1:nrow(mono_JUb19_raw), n_boot, replace = TRUE)
    mono_samples <- mono_JUb19_raw[mono_idx, ]
    
    # 隨機抽出 1000 次 Pairwise 的索引值
    pair_idx <- sample(1:nrow(pair_data), n_boot, replace = TRUE)
    pair_samples <- pair_data[pair_idx, ]
    tibble(
      boot_id = 1:n_boot,
      N_D_24_mono = mono_samples$M24, 
      r_D = mono_samples$r_D_raw,
      N_D_24_comp = pair_samples$D24,
      N_R_0_comp  = pair_samples$R0,
      interComp = ((N_D_24_comp-N_D_24_mono)/24)*(-1/N_R_0_comp)
    )
  }) %>%
  ungroup()


## 6 抽 1000 不做mean
interComp_boot_plot_indi_JUb19 <- ggplot(boot_fitness_df, 
                              aes(x = reorder(strain, interComp), y = interComp, fill = strain, color = strain)) +
  geom_violin(alpha = 0.3, color = NA, trim = FALSE) + 
  stat_summary(fun.data = "mean_cl_boot", geom = "errorbar", width = 0.3, size = 1, color = "black") +
  stat_summary(fun = "mean", geom = "point", size = 4, color = "black", shape = 18) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  
  scale_fill_manual(values = my_colors) +
  scale_color_manual(values = my_colors) +
  theme_minimal() +
  labs(title = "Interspecies Competition (1000x Bootstrap Simulation)",
       subtitle = "JUb19 as donor",
       x = "Bacterial recipient", y = "Interspecies competition") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

print(interComp_boot_plot_indi_JUb19)
```

#MSPm1
```{r}

# 提取 Donor (JUb19) 的單獨成長平均值 (Ki)
K_Donor_val <- monoMean_df %>%
  filter(strain == "MSPm1") %>%
  summarise(K_D_mean = mean_M24, r_D = (mean_logM24 - mean_logM0)/24 )

fitness_df_MSPm1 <- pairMSPm1 %>%
  # 合併 Recipient 的 K 值
  left_join(monoMean_df, by = "strain") %>%
  mutate(
    # 定義 Ki (Donor) 為剛才算出的常數
    N_D_24_mono = K_Donor_val$K_D_mean,
    N_D_24_comp = D24,
    N_D_0_comp = D0,
    N_R_0_comp = R0 ,
    r_D = K_Donor_val$r_D,
    # 計算 interspecies competition
    interComp = ((N_D_24_comp-N_D_24_mono)/24)*(-1/N_R_0_comp)
  )

fitness_df_MSPm1

interComp_plot_MSPm1 <- ggplot(fitness_df_MSPm1, 
                               aes(x = reorder(strain, interComp), y = interComp, color = strain)) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.3, size = 1, color = "black") +
  stat_summary(fun = "mean", geom = "point", size = 4, color = "black", shape = 18) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = my_colors) +
  theme_minimal() +
  labs(title = "Interspecies competition (Mean ± SE)",
       subtitle = "MSPm1 as donor",
       x = "Bacterial recipient", y = "Interspecies competition") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

print(interComp_plot_MSPm1)

# 6 抽 1000 不做mean
mono_MSPm1_raw <- mono_df %>%
  filter(strain == "MSPm1") %>%
  # 確認欄位沒有 NA，並計算出每次重複的 r_D
  mutate(r_D_raw = (logM24 - logM0) / 24)

# 設定 Bootstrap 抽樣次數
n_boot <- 1000

# 步驟 2：執行 Bootstrap 重抽樣
set.seed(42) 
boot_fitness_df <- pairMSPm1 %>%
  group_by(strain) %>%
  group_modify(~ {
    pair_data <- .x
    
    # 隨機抽出 1000 次 Mono 的索引值 (replace = TRUE 允許重複抽樣)
    mono_idx <- sample(1:nrow(mono_MSPm1_raw), n_boot, replace = TRUE)
    mono_samples <- mono_MSPm1_raw[mono_idx, ]
    pair_idx <- sample(1:nrow(pair_data), n_boot, replace = TRUE)
    pair_samples <- pair_data[pair_idx, ]
    tibble(
      boot_id = 1:n_boot,
      N_D_24_mono = mono_samples$M24, 
      r_D = mono_samples$r_D_raw,
      N_D_24_comp = pair_samples$D24,
      N_R_0_comp  = pair_samples$R0,
      interComp = ((N_D_24_comp-N_D_24_mono)/24)*(-1/N_R_0_comp)
    )
  }) %>%
  ungroup()



interComp_boot_plot_indi_MSPm1 <- ggplot(boot_fitness_df, 
                              aes(x = reorder(strain, interComp), y = interComp, fill = strain, color = strain)) +
  geom_violin(alpha = 0.3, color = NA, trim = FALSE) + 
  stat_summary(fun.data = "mean_cl_boot", geom = "errorbar", width = 0.3, size = 1, color = "black") +
  stat_summary(fun = "mean", geom = "point", size = 4, color = "black", shape = 18) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  
  scale_fill_manual(values = my_colors) +
  scale_color_manual(values = my_colors) +
  theme_minimal() +
  labs(title = "Interspecies Competition (1000x Bootstrap Simulation)",
       subtitle = "MSPm1 as donor",
       x = "Bacterial recipient", y = "Interspecies competition") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

print(interComp_boot_plot_indi_MSPm1)

```



